Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-07 sports_analytics.Rmd
outlined epv for sports with continuous states
edits after f3d76af; #5

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    "nflverse"
  ),
  file = "pkgs.bib"
)
```

@@

Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data. @cervone2016multiresolution introduces expected possession value (EPV), providing a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations.

@@

A related, but different, concept is the notion of *win probability*.
Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
Extending the mathematical framework we defined in Section \ref{sec:ev}, let $W$ be a binary random variable that indicates a win for a given team. 
Then, 
$$
  \Pr[W | s] \,,
$$
is the win probability in the state $s$.

This win probability is closely related to the expected value of a state. 
@Albert2015 defines the win probability as:
$$
  \Pr[W | s] = \sum_{X \geq 0} \Pr[X | s] \cdot \Pr[W | X, s] \,,
$$
where $\Pr[W | X, s]$ is the probability that the team will win the game given that they score $X$ points from state $s$.

Win probability is easily extended to provide a measure of the impact of sports plays and individual player contributions, as discussed in @Albert2015. 
Given its popularity, recent books on sports analytics often offers multiple chapters entirely on win probability. 
These include @Albert2001curve, @Schwarz2005numbers, @Tango2007book, @Albert2017handbook, and @Winston2022. 
In this section, we discuss notable previous work on win probability in baseball, football, basketball, and several other sports.

@@

@Albert2015 provided a historical overview of the use of win probability in baseball. This notion goes back to at least as early as @Lindsey1961, where the expected win probability after each inning was calculated based on the distribution of runs scored in each inning. @Mills1970 utilized win probability to introduce Player Win Average (PWA), a measure of a player's contribute to the game outcome. In particular, PWA was computed as 
$$
PWA = \frac{Win \ Points}{Win \ Points + Loss \ Points},
$$
where the win and loss points represent how much the player positively and negatively impact their team's probability of winning after each play.

@@

In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football. These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require little assumptions, and produce feature importance scores.

@Lock2014 used random forests as a means for estimating win probability before each play in a football game. Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down. In addition, @Yurko2019 estimated win probability in the NFL using a generalized additive model (GAM), as part of the `nflscrapr` package and nflWAR framework. This model takes into account the estimated expected points obtained from the model described in \ref{sec:ev}, along with other predictors for time, current half, and timeouts. The two win probability frameworks proposed by @Lock2014 and @Yurko2019 were also implemented in @Yam2019 with minimal modifications. Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision making in gridiron football.

A vital highlight of @Yurko2019's win probability model is that it is fully reproducible and uses publicly available data. One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work", which ultimately inspired \pkg{nflfastR}, known as the successor to `nflscrapr`. Figure \ref{fig:nflfastr_wp} shows a win probability graph for the 2017-18 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. The chance of winning for each team is obtained using the \pkg{nflfastR} R package, which implemented a gradient boosting model via the `xgboost` library for estimating win probabilities. Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively. The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 seconds left in the game. The last play of this game, famously known as the Minneapolis Miracle, resulted in a drastic swing in win probabilities for both teams. With 10 seconds remaining of game time, the Vikings begin the final possession with a 25.3% chance of winning. Their probability increased to a perfect 1 when Stefon Diggs scored a game-winning 61-yard receiving touchdown as the game clock expired.

@@

```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs Vikings in the 2017-18 NFL Playoffs."}
library(nflfastR)
load_pbp(2017) |> 
  filter(season_type == "POST",
         home_team == "MIN",
         away_team == "NO",
         !is.na(home_wp),
         !is.na(away_wp)) |>
  select(game_seconds_remaining, home_wp, away_wp) |> 
  pivot_longer(!game_seconds_remaining,
               names_to = "team",
               values_to = "wp") |>
  ggplot(aes(game_seconds_remaining, wp, color = team)) +
  geom_line(size = 1.5) +
  scale_x_reverse(breaks = seq(0, 3600, 900)) + 
  scale_color_manual(values = c(saints, vikings), 
                     labels = c("New Orleans Saints", "Minnesota Vikings")) +
  # geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  # geom_vline(xintercept = seq(0, 2700, 900), linetype = "dashed", color = "gray") + 
  labs(x = "Time Remaining (seconds)",
       y = "Win Probability",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

@@

@Stern1994 used Brownian motion to investigate the scoring process of basketball. Let $p(l, t)$ represents the win probability for the home team given an $l$-point lead after $t$ seconds of game time. The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. In particular, 
$$
p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
,
$$
Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.

On a related note, @Deshpande2016 extended @Stern1994's framework by applying it in a Bayesian setting. In particular, a Bayesian lasso regression model was proposed to assess the impact of individual players on their team's win probability at a given time of a basketball game. This model assumes independence of observation, constant variability in win probability, and also addresses multicollinearity.

@McFarlane2019 used logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA. The approach takes into account the remaining game time, score difference, and point spread. This basketball win probability model is then applied to the calculation of the End-of-game Tactics Metric (ETM), measuring how the win probability differs between the optimal and on-court actual decisions.

@@

The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team. @Brenzel2019 used three-dimensional Markov models to estimate win probability throughout a curling match. In particular, the authors proposed both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game. In esports, @Maymin2020 used logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. More recently, @Guan2022 developed an in-game win probability model for the National Rugby League using functional data analysis. In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.

@@

The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. For example, we consider the data given in @agresti2018introduction, page 247 on tennis results from 2014 to 2018 for five men's professional players Djokovic, Federer, Murray, Nadal, and Wawrinka. We fitted a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.

@@

Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is
$$
\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768
$$

@@

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978]. The Elo rating was originally developed for chess, but also has been used to estimate team strength in other sports. See @koning2017rating for more information on applications of the Elo rating in soccer.

@Glickman1998 proposed a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. In particular, this paper modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. See also @glickman2017estimating for more discussion on estimating team strengths in American football. More recently, @Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues. Betting moneylines were used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.

>>>>

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    "nflscrapr",
    "nflverse",
    "xgboost"
  ),
  file = "pkgs.bib"
)
```

@@

Even in sports where the concept of a state is not so easy to define, like basketball, the value of a possession can be estimated with player tracking data.

(a few sentences on tracking data)

@cervone2016multiresolution introduces expected possession value (EPV), providing a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations.

@@

A related, but different, concept is the notion of *win probability*.
Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
Extending the mathematical framework we defined in Section \ref{sec:ev}, let $W_i$ be a binary random variable that indicates a win for team $i$. 
Then, 
$$
  \Pr[W_i | s] \,,
$$
is the win probability for team $i$ in the state $s$.

This win probability is closely related to the expected value of a state. 
@Albert2015 defines the win probability as:
$$
  \Pr[W_i | s] = \sum_{X \geq 0} \Pr[X | s] \cdot \Pr[W_i | X, s] \,,
$$
where $\Pr[W_i | X, s]$ is the probability that team $i$ will win the game given that they score $X$ points from state $s$.

Win probability is easily extended to provide a measure of the impact of sports plays and individual player contributions, as discussed in @Albert2015. 
Given its popularity, recent books on sports analytics often dedicate multiple chapters entirely to win probability. 
These include @Albert2001curve, @Schwarz2005numbers, @Tango2007book, @Albert2017handbook, and @Winston2022.

In this section, we discuss notable previous work on win probability in baseball, football, basketball, and several other sports.

@@

@Albert2015 provides a historical overview of the use of win probability in baseball. 
This notion goes back to at least as early as @Lindsey1961, where the expected win probability after each inning was calculated based on the distribution of runs scored in each inning. 
@Mills1970 utilize win probability to introduce Player Win Average (PWA), a measure of a player's contribution to the game outcome. 
In particular, PWA is computed as 
$$
PWA = \frac{Win \ Points}{Win \ Points + Loss \ Points},
$$
where the win and loss points represent how much the player positively and negatively impact their team's probability of winning after each play.
In effect, the win points are the sum of the changes in $\Pr[W_i | s]$ from one state to the next.

@@

In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football. 
These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.

@Lock2014 use random forests as a means for estimating win probability before each play in a football game. 
Covariates included in this win probability model are the current down, score differential, time remaining, adjusted score, point spread, number of timeouts remaining for each team, total points scored, current yardline, and yards to go for a first down.

In addition, @Yurko2019 estimate win probability in the NFL using a generalized additive model (GAM), as part of the \pkg{nflscrapr} package (@R-nflscrapr) and nflWAR framework. 
This model takes into account the estimated expected points obtained from the model described in Section \ref{sec:ev}, along with other predictors for time, current half, and timeouts. 
The two win probability frameworks proposed by @Lock2014 and @Yurko2019 were also implemented in @Yam2019 with minimal modifications. Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in gridiron football.

A vital highlight of @Yurko2019's win probability model is that it is fully reproducible and uses publicly available data. 
One of @Yurko2019's goals was also to encourage researchers to "use, explore, and improve upon our work," which ultimately inspired \pkg{nflfastR} (@R-nflfastR), now considered the successor to \pkg{nflscrapr}.

Figure \ref{fig:nflfastr_wp} shows a win probability graph for the January 14, 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. 
We obtain the estimated probability of winning for each team using the \pkg{nflfastR} R package, which implements a gradient boosting model via the \pkg{xgboost} library (@R-xgboost) for estimating win probabilities. 
Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively. 
The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 left in the game. 
The last play of this game, famously known as the [Minneapolis Miracle](https://en.wikipedia.org/wiki/Minneapolis_Miracle), resulted in a drastic swing in win probabilities for both teams. 
With 10 seconds remaining in the game, the Vikings begin the final possession with a 25.3% chance of winning. 
Their probability increased to a perfect 1 when Stefon Diggs scored a game-winning 61-yard receiving touchdown as the game clock expired.

@@

```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs. Vikings in the 2017--18 NFL Playoffs."}
library(nflfastR)
x <- load_pbp(2017) |> 
  filter(
    season_type == "POST",
    home_team == "MIN",
    away_team == "NO",
    !is.na(home_wp),
    !is.na(away_wp)
  )
x |>
  select(game_seconds_remaining, home_wp, away_wp) |> 
  pivot_longer(
    !game_seconds_remaining,
    names_to = "team",
    values_to = "wp"
  ) |>
  ggplot(aes(game_seconds_remaining, wp, color = team)) +
  scale_x_reverse(breaks = seq(0, 3600, 900)) + 
  scale_color_manual(
    values = c(saints, vikings), 
    labels = c("New Orleans Saints", "Minnesota Vikings")
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(size = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

@@

@Stern1994 uses Brownian motion to investigate the scoring process of basketball. 
Let $p(l, t)$ represent the win probability for the home team given an $l$-point lead after $t$ seconds of game time. 
The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. 
In particular, 
$$
p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
,
$$
Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.

On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
In particular, they propose a Bayesian lasso regression model to assess the impact of individual players on their team's win probability at a given time of a basketball game. 
This model assumes independence of observation, constant variability in win probability, and also addresses multicollinearity.

@McFarlane2019 uses logistic regression to estimate win probability for evaluating end-of-game decisions in the NBA. 
The approach takes into account the remaining game time, score difference, and point spread. 
This basketball win probability model is then applied to the calculation of the End-of-game Tactics Metric (ETM), measuring how the win probability differs between the optimal and on-court actual decisions.

@@

The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team. 
@Brenzel2019 uses three-dimensional Markov models to estimate win probability throughout a curling match. 
In particular, the authors propose both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game. 
In esports, @Maymin2020 uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
More recently, @Guan2022 develops an in-game win probability model for the National Rugby League using functional data analysis. 
In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.

@@

The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
As an example, we consider the data given in @agresti2018introduction (page 247) on tennis results from 2014--2018 for five men's professional players Novak Djokovic, Roger Federer, Andy Murray, Rafael Nadal, and Stan Wawrinka. 
We fit a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.

@@

Table \ref{tab:tennisbtm} shows the estimated coefficients of the fitted BTM. 
According to the abilities, the players are ranked as follows: 1) Djokovic, 2) Federer, 3) Wawrinka, 4) Nadal, 5) Murray.
To obtain win probability, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
$$
\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768
$$

@@

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
See @koning2017rating for more information on applications of the Elo rating in soccer.

@Glickman1998 proposed a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
In particular, they modeled point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources variation in team strengths. 
See also @glickman2017estimating for more discussion on estimating team strengths in American football. 
More recently, @Lopez2018 extended @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
Betting moneylines were used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
